---
title: 'Modelling COVID-19 disease severity: just how dangerous is it for me?'
author: Daniel Lefever
date: '2020-07-30'
slug: modelling-covid-19-disease-severity-just-how-dangerous-is-it-for-me
categories:
  - COVID-19
  - disease severity
  - model
  - probability
  - R
  - statistics
  - epidemiology
tags:
  - COVID-19
  - model
  - probability
  - R
  - statistics
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

# Introduction

Lets forgot about the demographics and stick with a single IFR 0.65% for the moment. Since only 0.65% of people die, then that must mean 99.35% of people will ultimately recover. You can see^[Not that I think anyone needs help visualizing how large of a difference this is, I just like pictures.] in \@ref(fig:naive_binary_mortality) that the recovered dwarfs deaths, to the point where it's hard to even see the deaths at all. The issue here is that recovery can be misleading. When I think of recovery, I think of a return to baseline^[For example, when I get a cold, I'm miserable for anywhere from 3-14 days. After that, I'm back to normal. Sometimes, even better because of all the rest.]. That's unfortunately just not always the case with COVID-19. 

```{r naive_binary_mortality, fig.cap="Binary outcome"}
library(tidyverse)
library(ggplot2)
library(cowplot)
plot_data <- tibble(status=c("Recovered", "Dead"), value=c(.9935, .0065))

ggplot(data=plot_data, aes(x=status, y=value)) + 
  geom_col() + 
  scale_y_continuous(labels = scales::percent_format()) + 
  theme_cowplot() + 
  theme(text = element_text(face="bold")) + labs(x="", y="Percentage Infected with COVID-19")


```